---
title: "Accessing Web API Data Using R"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Accessing Web API Data Using R

### STEP 0 - Load the libraries
```{r, message=FALSE, warning=FALSE}
library(rgbif)      # -> for accessing GBIF data
library(sf)         # -> for spatial data manipulation
library(dplyr)      # -> for data manipulation
library(ggplot2)    # -> for plotting
library(rnaturalearth) # -> for spatial data access
```

### STEP 1 - Import the data
```{r, message=FALSE, warning=FALSE}
# STEP 1 - Make request to the GBIF API using rgbif
search_params <- list(
  scientificName = "Bubo scandiacus",
  country = "CA",
  stateProvince = "Quebec",
  year = 2010:2021,
  limit = 10
)

# Using rgbif to get occurrences
occ_data <- occ_search(
    scientificName = search_params$scientificName,
    stateProvince = search_params$stateProvince,
    eventDate = search_params$eventDate,
    limit = search_params$limit
)

# Extract data
data <- occ_data$data

# Print the data
print(data)
```

### STEP 2 - Data manipulation and cleaning

The `rgbif` package handles JSON parsing internally, so there's no need for manual parsing. The `occ_search` function retrieves data directly in a list format, which we transform into an `sf` object for spatial operations.

```{r, message=FALSE, warning=FALSE}
data <- data |>
  sf::st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326) |>
  sf::st_transform(32198)

# Data summary
print(data)
```

### STEP 3 - Visualize the data

The `ggplot2` package is used to plot spatial data stored in the `sf` object along with base map data for the province of Quebec.

```{r, message=FALSE, warning=FALSE}
# Get the province of Quebec geometry
quebec <- ne_states(country = "canada", returnclass = "sf") |>
  filter(name_en == "Quebec") |>
  sf::st_transform(32198)

# Plot the data
ggplot() +
  geom_sf(data = quebec) +
  geom_sf(data = data, color = "red", size = 2) +
  theme_minimal() +
  labs(title = "Occurrences of Bubo scandiacus in Quebec",
       x = "Longitude", y = "Latitude")
```
